---
header-includes: 
- \newcommand{\R}{\textnormal{\sffamily\bfseries R}}}
title: "R Notebook"
output:
  pdf_document: default
  html_notebook: default
---
```{r include=FALSE}
library(ggplot2)
library(ggpubr)
library(kableExtra)
```


```{r include=FALSE}
#Quantiles functions
#-------------------
q025=function(x){quantile(x,0.025)}
q975=function(x){quantile(x,0.975)}


#Calcolo intevalli di credibilità
#--------------------------------
Coverage<-function(fun){
  
  Filt<-Filtervalues(fun)
  mean<-Filt$mean
  sd<-Filt$sd
  vec<-NULL
  for(t in 1:length(mean)){
  if(mean[t]-1.96*sd[t]<truex[t] & truex[t]<mean[t]+1.96*sd[t]){vec[t]<-1}else{vec[t]<-0}
  }
  return(mean(vec))
}

#RMSE
#----
RMSE<-function(x,xhat){sqrt(mean((x - xhat)^2))}
```


\textbf{Implementation}

In this section we present a practical illustration of the algorithms discussed. For the sake of simplicity we use a random walk plus noise model, i.e. the most basic form of a linear Gaussian state-space model. 
\begin{align}
y_{t}|x_{t} & \sim N(x_{t},\sigma^{2}) \\
x_{t}|x_{t-1} & \sim N(x_{t-1},\tau^{2}) \\
x_{0} & \sim N(m_{0},C_{0})
\end{align}
As already mentioned before, in this case the filtering distribution can be computed in closed form solutions using the Kalman filter. However, this toy example will be used also to illustrate more involving filtering strategies described in this work. We believe indeed that it represents a useful starting point to understand the logic of the algorithms which may be eventually replicated when dealing with more complex models. \
The filtering strategy is applied to 50 simulated data. Figure XX shows the simulated true states sequence assuming as data generating process the Equation (2) with $\tau^{2}=1$ and simulated observed sequence process form Equation (1) with $\sigma^{2}=1$.
```{r echo=FALSE, fig.align='center', fig.cap="Simulated random walk plus noise model", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}

set.seed(1991)
dlm.sim = function(n,sigma,tau,x0){
  x    = rep(0,n)  #state process
  y    = rep(0,n)  #obs process
  x[1] = rnorm(1,x0,tau)
  y[1] = rnorm(1,x[1],sigma)
  for (t in 2:n){
    x[t] = rnorm(1,x[t-1],tau)
    y[t] = rnorm(1,x[t],sigma)
  }
  return(list(x=x,y=y))
}

n=50
tau2=1
sigma2=1
sigma=sqrt(sigma2)
tau=sqrt(tau2)
x0=0

sim<-dlm.sim(n,sigma,tau,x0)
y<-sim$y
x<-sim$x
truex<-x

timeframe<-c(1:length(y))
df<-data.frame(timeframe,y,x)

ggplot(df,aes(x=timeframe))+
  geom_line(aes(y=y, linetype="Observations"))+
  geom_line(aes(y=x, linetype="True States"))+
  coord_cartesian(ylim = c(-4, +12)) +
  scale_linetype_manual(name="",
                      values=c("Observations" = 1,
                               "True States" = 3))+
  labs(x="Time",
       y="")+
  #ggtitle("States and Observations")+
  theme_bw()+
  theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
        legend.background = element_rect(fill = "white", colour = "gray30")) +
  theme(plot.title = element_text(hjust = 0.5))
```
The Kalman Filter for this model is implemented as described below.
\begin{itemize}
\item Initialize $\theta_{0} \sim N(m_{0},C_{0})$
\item For $t=1,..N$
\begin{enumerate}
\item Compute the one-step-ahead state predictive distribution at time $t-1$
\begin{align*}
x_{t}|y_{1:t-1} & \sim N(a_{t},R_{t})\\
a_{t} & = m_{t-1}\\
R_{t} & = C_{t-1}+\tau^2
\end{align*}
\item Compute the filtering distribution at time $t$ as $p(x_{t}|y_{1:t}) \propto p(x_{t}|y_{1:t-1})p(y_{t}|x_{t})$, i.e. the product of the one-step-ahead state predictive distribution and the likelihood
\begin{align*}
x_{t}|y_{1:t} & \sim N(m_{t},C_{t}) \\
m_{t} & = \big(1-\frac{R_{t}}{R_{t}+\sigma^2}\big)a_{t}+\frac{R_{t}}{R_{t}+\sigma^2}y_{t} \\
C_{t} & = \frac{R_{t}}{R_{t}+\sigma^2}\sigma^2
\end{align*}
\end{enumerate}
\end{itemize}
These steps are summarize in the 
```{r}
DLM<-function(data,sig2,tau2,m0,C0){
  n  = length(data)
  m  = rep(0,n)
  C  = rep(0,n)
  for (t in 1:n){
    if (t==1){
      a = m0
      R = C0 + tau2
    }else{
      a = m[t-1]
      R = C[t-1] + tau2
    }
    A = R/(R+sig2)
    m[t] = (1-A)*a + A*y[t]
    C[t] = A*sig2
  }
  return(list(m=m,C=C))
}
```
In the Figure below we show the filtered states states estimated using Kalman Filter with $x_{0} \sim N(0,100)$ and  $\sigma^{2}=\tau^{2}=1$. The filtered states follow the observations closely and they provide a good approximation of the true states.
```{r echo=FALSE, fig.align='center', fig.cap="Kalman Filtered States with credible interval (in red)", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
#Kalman Filter
#-------------
m0=0
C0=100
filtval<-DLM(y,sigma2,tau2,m0,C0)
m<-filtval$m
C<-filtval$C
Low = m + qnorm(0.025)*sqrt(C)
Up = m + qnorm(0.975)*sqrt(C)

#Put everything in a data frate (for ggplot)
timeframe<-c(1:length(y))
DLM.df<-data.frame(timeframe,y,x,m,C,Low,Up)

#Plots (observetions in black and filtered states in red)
ggplot(DLM.df,aes(x=timeframe))+
  geom_line(aes(y=y, col="Observations", linetype="Observations"))+
  geom_line(aes(y=x, col="True States", linetype="True States"))+
  geom_line(aes(y=m, col="Filtered States", linetype="Filtered States"))+
  geom_ribbon(aes(ymin = Low, ymax = Up),
              fill="red",alpha=0.16) +
    scale_color_manual("",
                      values=c("Observations" = "black",
                               "True States" = "black",
                               "Filtered States"= "red"))+
  scale_linetype_manual("",
                      values=c("Observations" = 1,
                               "True States" = 3,
                               "Filtered States"=1))+
  labs(x="Time",
       y="")+
  ggtitle("Kalman Filter")+
  theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
        legend.background = element_rect(fill = "white", colour = "gray30")) +
  theme(plot.title = element_text(hjust = 0.5))
```

\textbf{Implementation}
With reference to the random walk plus noise of section XX, let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \frac{1}{N}\sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$. For $t=1,...,n$ where $n$ is the length of the sample, at any iteration
\begin{itemize}
\item Draw $x_{t}^{(i)} \sim N(x_{t-1}^{(i)},\tau^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t-1})$
\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t})$
\end{itemize}
```{r}
SISfun<-function(data,N,m0,C0,tau,sigma){
  xs<-NULL
  ws<-NULL
  ess<-NULL
  x  = rnorm(N,m0,sqrt(C0))
  w  = rep(1/N,N)
  for(t in 1:length(data)){
    x    = rnorm(N,x,tau)                   #sample from N(x_{t-1},tau)
    w    = w*dnorm(data[t],x,sigma)         #update weight
    xs = rbind(xs,x)
    ws = rbind(ws,w)
    
    wnorm= w/sum(w)                         #normalized weight
    ESS  = 1/sum(wnorm^2)                   #effective sample size
    
    ess =rbind(ess,ESS)
  }
  
  return(list(xs=xs,ws=ws,ess=ess))
}
```

```{r echo=FALSE}
N=1000
sis<-SISfun(y,N,m0,C0,tau,sigma)
sisxs<-sis$xs
sisws<-sis$ws
sisess<-sis$ess
sis.data<-data.frame(df,sisxs,sisws,sisess)

ggplot(sis.data,aes(x=timeframe))+
  geom_line(aes(y=sisess))+
    labs(x="Time",
       y="")+
  ggtitle("Effective Sample size")+
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5))
```

```{r echo=FALSE, fig.align='center', fig.cap="SIS Filtered States with credible interval (in red)", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
Filtervalues<-function(fun){
  xhat<-sapply(1:n,function(i)
    weighted.mean(fun$xs[i,],fun$ws[i,]))
  sdhat<-sapply(1:n,function(i)
    sqrt(weighted.mean((fun$xs[i,]-xhat[i])^2,fun$ws[i,])))
  
  return(list(mean=xhat,sd=sdhat))
  
}

Filtplot<-function(dataframe,fun,title){

Filt<-Filtervalues(fun)
mean<-Filt$mean
sd<-Filt$sd
dataframe<-data.frame(dataframe,mean,sd)
  
ggplot(dataframe,aes(x=timeframe))+
  geom_line(aes(y=y, col="Observations", linetype="Observations"))+
  geom_line(aes(y=x, col="True States", linetype="True States"))+
  geom_line(aes(y=mean, col="Filtered States", linetype="Filtered States"))+
  geom_ribbon(aes(ymin = mean-1.96*sd, ymax = mean+1.96*sd),
              fill="red",alpha=0.16) +
    scale_color_manual("",
                      values=c("Observations" = "black",
                               "True States" = "black",
                               "Filtered States"= "red"))+
  scale_linetype_manual("",
                      values=c("Observations" = 1,
                               "True States" = 3,
                               "Filtered States"=1))+
  labs(x="Time",
       y="")+
  ggtitle(title)+
  theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
        legend.background = element_rect(fill = "white", colour = "gray30")) +
  theme(plot.title = element_text(hjust = 0.5))
}

Filtplot(df,sis,title="Sequential Importance Sampling Filter")

```
\
\textbf{Particle Filter}
\
With reference to the random walk plus noise of section XX, let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \frac{1}{N}\sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$. For $t=1,...,n$ where $n$ is the length of the sample, at any iteration
\begin{itemize}
\item Draw $x_{t}^{(i)} \sim N(x_{t-1}^{(i)},\tau^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t-1})$
\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t})$
\end{itemize}
In addition, when $ESS<ESS_{0}$ \footnote{In our example we fix $ESS_{0}=N/2$, this is an arbitrary common rule of thumb.}, resempling applies
\begin{itemize}
\item Draw a sample of size N, $x_{t}^{(1)},...,x_{t}^{(N)}$,from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
\end{itemize}
```{r}
PFfun<-function(data,N,m0,C0,tau,sigma,r){
  if(missing(r)){r=2}else{}
  xs<-NULL
  ws<-NULL
  ess<-NULL
  x  = rnorm(N,m0,sqrt(C0))
  w  = rep(1/N,N)
   
  for(t in 1:length(data)){
    
    x<-rnorm(N,x,tau)
    w1<-w*dnorm(data[t],x,sigma)
    
    w = w1/sum(w1)
    ESS  = 1/sum(w^2)
    
    if(ESS<(N/r)){
      index<-sample(N,size=N,replace=T,prob=w)
      x<-x[index]
      w<-rep(1/N,N)
    }else{}
    
    xs = rbind(xs,x)
    ws = rbind(ws,w)
    ess =rbind(ess,ESS)
  }
  return(list(xs=xs,ws=ws,ess=ess))
}
```

```{r echo=FALSE, fig.align='center', fig.cap="Effective Sample Size", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
N=1000

pf<-PFfun(y,N,m0,C0,tau,sigma)

ESSplot<-function(dataframe,fun,threshold){
if(missing(threshold)){threshold=T}else{}
xs<-fun$xs
ws<-fun$ws
ess<-fun$ess
dt<-data.frame(dataframe,xs,ws,ess)

ggplot(dt,aes(x=timeframe))+
  geom_line(aes(y=ess))+
  geom_line(aes(y=500),col="orange")+
  labs(x="Time",
       y="")+
  ggtitle("Effective Sample Size")+
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5))
}
ESSplot(df,pf)

```

```{r echo=FALSE, fig.align='center', fig.cap="Particle Filtered States with credible interval (in red)", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
Filtplot(df,pf,"Particle Filter")
```

\textbf{Particle Filter Petroni more efficient}
```{r}
PF2fun<-function(data,N,m0,C0,tau,sigma,r){
  if(missing(r)){r=2}else{}
  xs<-NULL
  ws<-NULL
  ess<-NULL
  x  = rnorm(N,m0,sqrt(C0))
  importancesd<-sqrt(tau - tau^2 /(tau + sigma))
  predsd <- sqrt(sigma+tau)
  w  = rep(1/N,N)
  
  for(t in 1:length(data)){
    
    means<-x+(tau/(tau+sigma))*(data[t]-x)
    x<-rnorm(N,means,importancesd)
    w1<-w*dnorm(data[t],x,predsd)
    
    w = w1/sum(w1)
    ESS  = 1/sum(w^2)
    
    if(ESS<(N/r)){
      index<-sample(N,size=N,replace=T,prob=w)
      x<-x[index]
      w<-rep(1/N,N)
    }else{}
    
    xs = rbind(xs,x)
    ws = rbind(ws,w)
    ess =rbind(ess,ESS)
  }
  return(list(xs=xs,ws=ws,ess=ess))
}
```

\textbf{comparison between Particle Filters}

```{r echo=FALSE}
pf1<-PFfun(y,1000,0,100,1,1)
pf2<-PF2fun(y,1000,0,100,1,1)
plotpf1<-Filtplot(df,pf1,"transition")
plotpf2<-Filtplot(df,pf2,"more efficient")

ggarrange(plotpf1,plotpf2)
```

```{r}
Comparethetwo<-matrix(NA,nrow=2,ncol=2)
colnames(Comparethetwo)<-c("Coverage","RMSE")
rownames(Comparethetwo)<-c("transition","more efficient")
Comparethetwo[1,1]<-Coverage(pf1)
Comparethetwo[2,1]<-Coverage(pf2)
Comparethetwo[1,2]<-RMSE(truex,Filtervalues(pf1)$mean)
Comparethetwo[2,2]<-RMSE(truex,Filtervalues(pf2)$mean)
```

```{r echo=FALSE}
kbl(Comparethetwo, align="c", longtable = T, booktabs = T, caption="Comparison",digits=3)
```






\textbf{Guided Particle Filter}

Let's consider another toy example. Suppose that we have the following model

```{r}
GPFfun<-function(data,N,m0,C0,tau,sigma,r){
 if(missing(r)){r=2}else{}
   xs<-NULL
  ws<-NULL
  ess<-NULL
  x  = rnorm(N,m0,sqrt(C0))
  w  = rep(1/N,N)
  
  for(t in 1:length(data)){
    
    xprev<-x
    x<-rnorm(N,x,tau)
    w1<-w*dnorm(data[t],x,sigma)*dnorm(x,xprev,tau)*I(x>0)/dnorm(x,xprev,tau)
    
    w = w1/sum(w1)
    ESS  = 1/sum(w^2)
    
    if(ESS<(N/r)){
      index<-sample(N,size=N,replace=T,prob=w)
      x<-x[index]
      w<-rep(1/N,N)
    }else{}
    
    xs = rbind(xs,x)
    ws = rbind(ws,w)
    ess =rbind(ess,ESS)
  }
  return(list(xs=xs,ws=ws,ess=ess))
}
```


\textbf{}

With reference to the linear gaussian model of section XX, a very basic auxiliary particle sampling technique is 
\begin{itemize}
\item Let $\{(x_{t-1},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t-1}|y_{t-1})$
\end{itemize}
For $k=1,...,N$
\begin{itemize}
\item Draw $I_{k}$ with $P(I_{k}) \propto w_{t-1}^{(i)}f(y_{t}|g(x_{t-1}^{(i)})) $ where $g(x_{t-1}^{(i)})=E(X_{t}|X_{t-1})$
\item Draw $x_{t}^{(k)} \sim N(x_{t-1}^{(I_{k})},\tau^2)$
\item Set  $w_{t}^{(k)} = \frac{f_{N}(y_{t}|x_{t}^{(k)})}{f_{N}(y_{t}|g(x_{t-1}^{(I_{k})}))}$
\end{itemize}
The remaining steps are the same of the particle filter already seen in section XX.

```{r}
APFfun<-function(data,N,m0,C0,tau,sigma,r){
  if(missing(r)){r=2}else{}
  xs<-NULL
  ws<-NULL
  ess<-NULL
  x  = rnorm(N,m0,sqrt(C0))
  w  = rep(1/N,N)
  
  for(t in 1:length(data)){
    
    weight = w*dnorm(data[t],x,sigma)
    k   = sample(1:N,size=N,replace=TRUE,prob=weight)
    x1   = rnorm(N,x[k],tau)
    lw  = dnorm(data[t],x1,sigma,log=TRUE)-dnorm(data[t],x[k],sigma,log=TRUE)
    w   = exp(lw)
    w   = w/sum(w)
    ESS  = 1/sum(w^2)
    
    if(ESS<(N/r)){
      index<-sample(N,size=N,replace=T,prob=w)
      x1<-x1[index]
      w<-rep(1/N,N)
    }else{}
    
    x <- x1
    xs = rbind(xs,x)
    ws = rbind(ws,w)
    ess =rbind(ess,ESS)
    
  }
  return(list(xs=xs,ws=ws,ess=ess))
}
```

```{r echo=FALSE}
apf<-APFfun(y,N,m0,C0,tau,sigma)
ESSplot(df,apf)
```

```{r echo=FALSE, fig.align='center', fig.cap="Particle Filtered States with credible interval (in red)", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
Filtplot(df,apf,"Particle Filter")
```
\
\textbf{Liu and West Filter}
\
Consider the toy example of section XX. Let $\psi=(\sigma^{2},\tau^{2})$ be unknown and assign a gamma prior on these parameters
\begin{align*}
\sigma^{2}  & \sim G(\alpha_{v},\beta_{v}) \\
\tau^{2}  & \sim G(\alpha_{w},\beta_{w})
\end{align*}
or let assign them a uniform prior if we have no knowledge on hyperparameters.
After having drown the hyperparameters indipendently form their priors and having set $w_{0}^{(i)}=N^{-1}$, $i=1,...,N$, and $\hat{\pi}_{0}=\sum_{i=1}^{N}w_{0}^{(i)}\delta_{(x_{0}^{(i)},\psi^{(i)})}$, for t=1,..T 
\begin{itemize}
\item Compute $\hat{\psi}=E_{\hat{\pi}_{t-1}}(\psi)$ and $\Sigma=Var_{\hat{\pi}_{t-1}}(\psi)$. For $i=1,...,N$ set
\begin{align*}
m(\psi^{(i)}) & =a\psi^{(i)}+(1-a)\overline{\psi}\\
v(\psi^{(i)}) & =(1-a^2)\Sigma
\end{align*}
and
\begin{align*}
\alpha(\psi^{(i)}) & =\frac{m(\psi^{(i)})^2}{v(\psi^{(i)})} \\
\beta(\psi^{(i)}) & =\frac{m(\psi^{(i)})}{v(\psi^{(i)})}
\end{align*}
For $k=1,...,N$
\begin{itemize}
\item Draw $I_{k}$ with $P(I_{k}=i) \propto w_{t-1^{(i)}}f_{N}(y_{t}|g(x_{t}^{(i)}),m(\psi^{(i)}))$ where for simplicity $g(x_{t}^{(i)})=E(x_{t}|x_{t-1},m(\psi^{(i)}))$
\item Draw $\psi^{(k)} \sim G(\alpha(\psi^{(I_{k})}),\beta(\psi^{(I_{k})}))$
\item Draw $x_{t}^{(k)} \sim N(x_{t-1}^{(I_{k})},\tau^{{2}^{(k)}})$
\item Set $\tilde{w}_{t}^{k}=\frac{f_N(y_{t}|x_{t}^{(k)},\psi=\psi^{(k)})}{f_N(y_{t}|g(x_{t}^{(I_{k})}),\psi=m(\psi)^{(I_k)})}$
\end{itemize}
\item Normalize the weights
\item Compute the effective sample size ($ESS$)
\item If $ESS<N/2$, resample:
\begin{itemize}
\item Draw a sample of size N, $x_{t}^{(1)},...,x_{t}^{(N)}$,from the discrete distribution $P((x_{t},\psi)=(x_{t}^{(i)},\psi^{(i)}))=w_{t}^{(i)},\ \ i=1,...,N$
\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
\end{itemize}

\end{itemize}
```{r}
LWfun<-function(data,N,m0,C0,alphav,betav,alphaw,betaw,delta,unif,r){
  if(missing(r)){r=2}else{}
  xs     = rnorm(N,m0,sqrt(C0))
  if(unif==T){
  pars   = cbind(runif(N,0,10),runif(N,0,10))}else{}
  pars   = cbind(rgamma(N,shape=alphav,scale=betav),rgamma(N,shape=alphaw,scale=betaw))
  a      = (3*delta-1)/(2*delta)
  h2     = 1-a^2
  parss  = array(0,c(N,2,n))
  xss    = NULL
  ws     = NULL
  ess    = NULL
  w      = rep(1/N,N)
  for (t in 1:length(data)){
    meanV = weighted.mean(pars[,1],w)
    varV  = weighted.mean((pars[,1]-meanV)^2,w)
    meanW = weighted.mean(pars[,2],w)
    varW  = weighted.mean((pars[,2]-meanW)^2,w)
    
    muV = a*pars[,1]+(1-a)*meanV
    sigma2V = (1-a^2)*varV
    alphaV = muV^2/sigma2V
    betaV = muV/sigma2V
    
    muW = a*pars[,1]+(1-a)*meanW
    sigma2W = (1-a^2)*varW
    alphaW = muW^2/sigma2W
    betaW = muW/sigma2W
    
    weight      = w*dnorm(data[t],xs,sqrt(muV))
    k           = sample(1:N,size=N,replace=T,prob=weight)
    
    pars[,1]<-rgamma(N,shape=alphaV[k],rate=betaV[k])
    pars[,2]<-rgamma(N,shape=alphaW[k],rate=betaW[k])
    
    xsprevious<-xs[k]
    xs = rnorm(N,xs[k],sqrt(pars[,2]))
    
    w           = exp(dnorm( data[t],xs,sqrt(pars[,1]),log=T)-
                        dnorm( data[t],xsprevious,sqrt(muV[k]),log=T))
    w           = w/sum(w)
    ESS         = 1/sum(w^2)
    
    if(ESS<(N/r)){
      index<-sample(N,size=N,replace=T,prob=w)
      xs<-xs[index]
      pars<-pars[index,]
      w<-rep(1/N,N)
    }else{
      xs<-xs
      pars<-pars
    }
    
    
    xss         = rbind(xss,xs)
    parss[,,t]  = pars 
    ws          = rbind(ws,w)
    ess         = rbind(ess,ESS)
  }
  return(list(xss=xss,parss=parss,ws=ws,ess=ess))
}
```

```{r echo=FALSE}
delta=0.7
lwf<-LWfun(y,N,m0,C0,1,1,1,1,delta,unif=T)
ESSplot(df,lwf)
```

```{r echo=FALSE, fig.align='center', fig.cap="Particle Filtered States with credible interval (in red)", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
Filtplot(df,lwf,"Liu and West Filter")
```



# COMPARISON (ANOTHER SECTION)

```{r echo=FALSE, fig.align='center', fig.cap="Variances", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}


Kalmanvarplot<-ggplot(DLM.df,aes(x=timeframe))+
  geom_line(aes(y=C))+
  labs(x="Time",
       y="")+
  ggtitle("Kalman Filter")+
  theme_bw()


Varplot<-function(dataframe,fun,title){

Filt<-Filtervalues(fun)
mean<-Filt$mean
sd<-Filt$sd
dataframe<-data.frame(dataframe,mean,sd)

ggplot(dataframe,aes(x=timeframe))+
geom_line(aes(y=sd))+
  labs(x="Time",
       y="")+
  ggtitle(title)+
  theme_bw()

}


Varsisplot<-Varplot(df,sis,"SIS")
Varpfplot<-Varplot(df,pf,"PF")
Varapfplot<-Varplot(df,pf,"APF")
Varlwplot<-Varplot(df,pf,"LW")

ggarrange(Kalmanvarplot,Varsisplot,Varpfplot,Varapfplot,Varlwplot)

```

```{r include=FALSE}
truex<-x
estimatedx<-matrix(NA,ncol=6,nrow=6)
colnames(estimatedx)<-c("N","Threshold","KF","PF","APF","LWF")
estimatedx[,1]<-c(20,100,500,1000,1000,1000)
estimatedx[,2]<-c(0.5,0.5,0.5,0.5,0.25,0.1)


i=1
for(N in c(100,1000,10000)){

DLM1<-DLM(y,sigma2,tau2,m0,C0)
pf1<-PFfun(y,N,m0,C0,tau,sigma)
apf1<-APFfun(y,N,m0,C0,tau,sigma)
lwf1<-LWfun(y,N,m0,C0,1,1,1,1,delta,unif=T)

estimatedx[i,3]<-RMSE(truex,DLM1$m)
estimatedx[i,4]<-RMSE(truex,Filtervalues(pf1)$mean)
estimatedx[i,5]<-RMSE(truex,Filtervalues(apf1)$mean)
estimatedx[i,6]<-RMSE(truex,Filtervalues(lwf1)$mean)
i=i+1

}

i=4
for(k in c(2,5,10)){
  N=1000
DLM2<-DLM(y,sigma2,tau2,m0,C0)
pf2<-PFfun(y,N,m0,C0,tau,sigma,r=k)
apf2<-APFfun(y,N,m0,C0,tau,sigma,r=k)
lwf2<-LWfun(y,N,m0,C0,1,1,1,1,delta,unif=T,r=k)

estimatedx[i,3]<-RMSE(truex,DLM2$m)
estimatedx[i,4]<-RMSE(truex,Filtervalues(pf2)$mean)
estimatedx[i,5]<-RMSE(truex,Filtervalues(apf2)$mean)
estimatedx[i,6]<-RMSE(truex,Filtervalues(lwf2)$mean)
i=i+1


}



```

```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
kbl(estimatedx[1:3,], align="c", longtable = T, booktabs = T, caption="RMSE",digits=3)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
kbl(estimatedx[4:6,], align="c", longtable = T, booktabs = T, caption="RMSE",digits=3)
```

# STOCHASTIC VOLATILITY

Basic specification of Stochastic Volatilty Model
\begin{align}
y_{t}|x_{t} & \sim N(0,e^{x_{t}}) \\
x_{t}|x_{t-1} & \sim N(\alpha+\beta x_{t-1},\tau^2)
\end{align}
\
```{r include=FALSE}
#############################
#
#   y(t) = exp(h(t)/2)*v(t)
#   x(t) = alpha + beta*x(t-1) + tau*w(t)
#
#   v(t) ~ N(0,1)
#   w(t) ~ N(0,1)
#   x(0) ~ N(0,100)
#
#   PRIOR ON alpha,beta,tau2

# Simulating the data
# -------------------
set.seed(12345)
n     =  50
alpha = -0.0031
beta  =  0.9951
tau2  =  0.0074
tau   = sqrt(tau2)
y     = rep(0,n)
x     = rep(0,n)
x[1]  = alpha/(1-beta)
y[1]  = rnorm(1,0,exp(x[1]/2))
for (t in 2:n){
  x[t] = alpha + beta*x[t-1] + rnorm(1,0,tau)
  y[t] = rnorm(1,0,exp(x[t]/2))
}
alpha.true = alpha
beta.true  = beta
tau2.true  = tau2

timeframe<-c(1:length(y))
dfsv<-data.frame(timeframe,y,x)
```

```{r echo=FALSE, fig.align='center', fig.cap="SV", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
n=50
tau2=1
sigma2=1
sigma=sqrt(sigma2)
tau=sqrt(tau2)
x0=0

ggplot(dfsv,aes(x=timeframe))+
  geom_line(aes(y=y, linetype="Observations"))+
  geom_line(aes(y=x, linetype="True States"))+
  scale_linetype_manual(name="",
                      values=c("Observations" = 1,
                               "True States" = 3))+
  labs(x="Time",
       y="")+
  ggtitle("States and Observations")+
  theme_bw()+
  theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
        legend.background = element_rect(fill = "white", colour = "gray30")) +
  theme(plot.title = element_text(hjust = 0.5))
```

\textbf{Particle Filter}
\
```{r}
SVPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
  if(missing(r)){r=2}else{}
  xs<-NULL
  ws<-NULL
  ess<-NULL
  x  = rnorm(N,m0,sqrt(C0))
  w  = rep(1/N,N)
   
  for(t in 1:length(data)){
    
    x<-rnorm(N,alpha+beta*x,tau)
    w1<-w*dnorm(data[t],0,exp(x/2))
    
    w = w1/sum(w1)
    ESS  = 1/sum(w^2)
    
    if(ESS<(N/r)){
      index<-sample(N,size=N,replace=T,prob=w)
      x<-x[index]
      w<-rep(1/N,N)
    }else{}
    
    xs = rbind(xs,x)
    ws = rbind(ws,w)
    ess =rbind(ess,ESS)
  }
  return(list(xs=xs,ws=ws,ess=ess))
}
```

\textbf{Auxiliary Particle Filter}

```{r}
SVAPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
  if(missing(r)){r=2}else{}
  xs<-NULL
  ws<-NULL
  ess<-NULL
  x  = rnorm(N,m0,sqrt(C0))
  w  = rep(1/N,N)
  
  for(t in 1:length(data)){
    
    weight = w*dnorm(data[t],0,exp(x/2))
    k   = sample(1:N,size=N,replace=TRUE,prob=weight)
    x1   = rnorm(N,alpha+beta*x[k],tau)
    lw  = dnorm(data[t],0,exp(x/2),log=TRUE)-dnorm(data[t],0,exp(x/2),log=TRUE)
    w   = exp(lw)
    w   = w/sum(w)
    ESS  = 1/sum(w^2)
    
    if(ESS<(N/r)){
      index<-sample(N,size=N,replace=T,prob=w)
      x1<-x1[index]
      w<-rep(1/N,N)
    }else{}
    
    x <- x1
    xs = rbind(xs,x)
    ws = rbind(ws,w)
    ess =rbind(ess,ESS)
    
  }
  return(list(xs=xs,ws=ws,ess=ess))
}
```

\textbf{Liu and West}

```{r include=FALSE}
# Data and prior hyperparameters 
# ------------------------------
m0      = 0.0
C0      = 0.1
sC0     = sqrt(C0)
ealpha  = alpha
valpha  = 0.01
ebeta    = beta
vbeta    = 0.01
nu      = 3
lambda  = tau2
```

```{r}
SVLWfun<-function(data,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda){

xs = rnorm(N,m0,sqrt(C0))
pars   = cbind(rnorm(N,ealpha,sqrt(valpha)),rnorm(N,ebeta,sqrt(vbeta)),
               log(1/rgamma(N,nu/2,nu*lambda/2)))
delta  = 0.75
a      = (3*delta-1)/(2*delta)
h2     = 1-a^2
parss  = array(0,c(N,3,n))
xss    = NULL
ws     = NULL
ESS    = NULL
w      = rep(1/N,N)
for (t in 1:length(data)){
  
  #mpar        = apply(pars,2,mean) #alternatively use this 
  mpar<-c()
  for(i in 1:3){
  mpar[i]    = weighted.mean(pars[,i],w)
  }
  
  vpar        = var(pars)
  ms          = a*pars+(1-a)*matrix(mpar,N,3,byrow=T)
  mus         = pars[,1]+pars[,2]*xs 
  weight      = w*dnorm(data[t],0,exp(mus/2))
  k           = sample(1:N,size=N,replace=T,prob=weight)
  ms1         = ms[k,] + matrix(rnorm(3*N),N,3)%*%chol(h2*vpar)
  xt          = rnorm(N,ms1[,1]+ms1[,2]*xs[k],exp(ms1[,3]/2))
  w           = dnorm(data[t],0,exp(xt/2))/dnorm(data[t],0,exp(mus[k]/2))
  w           = w/sum(w)
  ESS         = 1/sum(w^2)
  
  
  if(ESS<(N/2)){
    index<-sample(N,size=N,replace=T,prob=w)
    xs<-xt[index]
    pars<-ms1[index,]
    w<-rep(1/N,N)
  }else{
    xs<-xt
    pars<-ms1
  }
  
  xss         = rbind(xss,xs)
  parss[,,t]  = pars 
  ws          = rbind(ws,w)
}
return(list(xs=xss,pars=parss,ws=ws))
}
```


```{r echo=FALSE, fig.align='center', fig.cap="Kalman Filtered States with credible interval (in red)", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
#Kalman Filter
#-------------
m0=0
C0=100
filtval<-DLM(y,sigma2,tau2,m0,C0)
m<-filtval$m
C<-filtval$C
Low = m + qnorm(0.025)*sqrt(C)
Up = m + qnorm(0.975)*sqrt(C)

#Put everything in a data frate (for ggplot)
timeframe<-c(1:length(y))
DLM.df<-data.frame(timeframe,y,x,m,C,Low,Up)

#Plots (observetions in black and filtered states in red)
plot0<-ggplot(DLM.df,aes(x=timeframe))+
  geom_line(aes(y=y, col="Observations", linetype="Observations"))+
  geom_line(aes(y=x, col="True States", linetype="True States"))+
  geom_line(aes(y=m, col="Filtered States", linetype="Filtered States"))+
  geom_ribbon(aes(ymin = Low, ymax = Up),
              fill="red",alpha=0.16) +
    scale_color_manual("",
                      values=c("Observations" = "black",
                               "True States" = "black",
                               "Filtered States"= "red"))+
  scale_linetype_manual("",
                      values=c("Observations" = 1,
                               "True States" = 3,
                               "Filtered States"=1))+
    coord_cartesian(ylim=c(-6,6))+
  labs(x="Time",
       y="")+
  ggtitle("Kalman Filter")+
  theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
        legend.background = element_rect(fill = "white", colour = "gray30")) +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r include=FALSE}
set.seed(12345)
N=5000
svpf<-SVPFfun(y,N,m0,C0,alpha,beta,tau)
svapf<-SVAPFfun(y,N,m0,C0,alpha,beta,tau)
svlw<-SVLWfun(y,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda)
```

```{r echo=FALSE, fig.align='center', fig.cap="Comparison filter performances", fig.height=8, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
plot1<-Filtplot(dfsv,svpf,"Particle Filter")
plot2<-Filtplot(dfsv,svapf,"Auxiliary Particle Filter")
plot3<-Filtplot(dfsv,svlw,"Liu e West Filter")

ggarrange(plot0,plot1,plot2,plot3)
```





